### 2013-03-28 doc/sosp13/memory.tex
minor edits

\section{Memory}

Applications written for Grappa utilize two forms of memory: local and global.
Local memory is local to a single core within a node in the system. Accesses
occur through conventional pointers. The compiler emits an access and the
memory is manipulated directly. Applications use local accesses for a number
of things in Grappa: the stack associated with a task, accesses to localized
global memory in caches (see below), and accesses to debugging infrastructure
that is local to each system node. Local pointers cannot access memory on
other cores, and are valid only on their home core.

Large data that is expected to be shared and accessed with low locality is
stored in Grappa's global memory. All global data must be accessed through
calls into Grappa's API, shown in Figure~\ref{fig:accessing-memory}.

\paragraph{Global memory addressing} Grappa provides two methods for storing
data in the global memory. The first is a distributed heap striped across all
the machines in the system in a block cyclic fashion. The
\texttt{global\_malloc} and \texttt{global\_free} calls are used to allocate
and deallocate memory in the global heap. Addresses to memory in the global
heap use \textbf{linear addresses}. Choosing the block size involves trading
off sequential bandwidth against aggregate random access bandwidth. Smaller
block sizes help spread data across all the memory controllers in the cluster,
but larger block sizes allow the locality-optimized memory controllers to
provide increased sequential bandwidth. The block size, which is configurable,
is typically set to 64 bytes, or the size of a single hardware cache line, in
order to exploit spatial locality when available.

%%% is the below important? -Luis
\comment{The heap metadata is stored on a single node. Currently all heap
allocation and free operations are serialized through this node; while this
has been sufficient for our benchmarks, it is easy to relax in the future
Grappa will provide parallel performance through
combining~\cite{MAMA,flatcombining}.}

Grappa also allows any local data on a core's stacks or heap to be
exported to the global address space to be made accessible to other
cores across the system. Addresses to global memory allocated in this
way use \textbf{2D global addresses}.  This uses a traditional PGAS
addressing model, where each address is a tuple of a rank in the job (or
global process ID) and an address in that process. The lower 48 bits of
the address hold a virtual address in the process. The top bit is set to
indicates that the reference is a 2D address (as opposed to linear
address). This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^{15}$ endpoints. Any node-local data can be made
accessible by other nodes in the system by wrapping the address and node
ID into a 2D global address. This address can then be accessed with a
delegate operation and even be cached by other nodes. The
address is converted at the destination into a canonical x86 address by replacing the upper
bits with the sign-extended upper bit of the virtual address. 2D
addresses may refer to memory allocated from a single processes' heap or
from a task's stack. Figure~\ref{fig:memory-structure} shows how 2D and
linear addresses can refer to other cores' memory.

\begin{figure}[htbp]
  \begin{center}
    \begin{minipage}{\columnwidth}
	\small
	\begin{tabular}{l}
      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
      	\texttt{\scriptsize global\_free( global\_address )} \\ 
      	Allocation in the global heap \\ \\
      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_read\_async( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_write\_async( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ 
      	Memory operations at the home core of a global address \\ \\
      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ 
		Perform cache operations to acquire/release global data.  \\
		Acquire copies all data to local node and returns a pointer. \\ 	
		Release optionally copies data back to global memory. \\
	\end{tabular}
      \caption{\label{fig:accessing-memory} Grappa API for memory accesses. \TODO{Update table with new calls. And async calls for delegates.}}     \end{minipage}
  \end{center}
\end{figure}

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:memory-structure} Global memory referencing in Grappa}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\paragraph{Global memory access} There are two general methods Grappa
applications use to {\em access} global memory. One method is to use {\em
delegate} operations, which are short memory accesses performed at the memory
location's home node. This method is used when there is no locality in the
access. The other method is to do {\em explicit caching} of memory regions,
used when the programmer expects that the associated computation has data
locality. Below we explain each in detail.

@@

Delegate operations can be either {\em synchronous} or {\em asynchronous}.
With synchronous operations, the task issuing the delegate call blocks until
the delegate operation completes, which is necessary, for example, to ensure
that synchronization has finished before continuing. Synchronization in Grappa
is implemented using synchronous delegate operations. Often remote data
accesses do not need to be performed synchronously. In order to avoid
unnecessary waiting, we support asynchronous delegate operations. For reads,
we support a ``futures''-like mechanism. Tasks may issue delegate reads in
parallel and block on the ``promises'' returned by asynchronous delegate
invocations. Delegate write operations may also be performed asynchronously,
and since they do not return data, they need a mechanism to detect completion.
Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
outstanding asynchronous operations of the local core. The code and can probe
the state of this object to determine when all outstanding requests are
complete. This is necessary to, say, make sure all outstanding operations are
complete before executing a synchronization operation.

@@

\vspace{1ex}
\textit{Explicit caching.} Grappa provides an API to fetch a global pointer of
any length and return a local pointer to a cached copy of the global memory.
Grappa cache operations have the usual read-only and read-write variants,
along with a write-only variant used to initialize data structures. Languages
for distributed shared memory systems have done optimizations to achieve a
similar goal. For example, the UPC compiler coalesces struct and array
accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's
message vectorization hoists small messages out of a loop
\cite{FortranD:1992}. Caching in Grappa additionally provides a mechanism for
exploiting temporal locality when available by operating on the data locally.

Under the hood, Grappa performs the mechanics of gathering chunks of data from
multiple system nodes and presenting a conventional appearing linear block of
memory as a local pointer into a cache. The strategy employed is to issue all
the constituent requests of the desired block (as Active Messages) and then
yield until all responses have occurred. Currently, Grappa caches are
\emph{not\/} kept coherent automatically. If that is desired, the programmer is
responsible for ensuring sufficient synchronization (via delegate operations)
occurs before acquiring or releasing cache objects.

>>>>

\section{Distributed Shared Memory}

Applications written for \Grappa utilize two forms of memory: local and global.
Local memory is local to a single core within a node in the system. Accesses
occur through conventional pointers. The compiler emits an access and the
memory is manipulated directly. Applications use local accesses for a number
of things in \Grappa: the stack associated with a task, accesses to localized
global memory in caches (see below), and accesses to debugging infrastructure
that is local to each system node. Local pointers cannot access memory on
other cores, and are valid only on their home core.

Large data that is expected to be shared and accessed with low locality is
stored in \Grappa's global memory. All global data must be accessed through
calls into \Grappa's API, shown in Figure~\ref{fig:accessing-memory}.

\paragraph{Global memory addressing} \Grappa provides two methods for storing
data in the global memory. The first is a distributed heap striped across all
the machines in the system in a block cyclic fashion. The
\texttt{global\_malloc} and \texttt{global\_free} calls are used to allocate
and deallocate memory in the global heap. Addresses to memory in the global
heap use \textbf{linear addresses}. Choosing the block size involves trading
off sequential bandwidth against aggregate random access bandwidth. Smaller
block sizes help spread data across all the memory controllers in the cluster,
but larger block sizes allow the locality-optimized memory controllers to
provide increased sequential bandwidth. The block size, which is configurable,
is typically set to 64 bytes, or the size of a single hardware cache line, in
order to exploit spatial locality when available.

%%% is the below important? -Luis
\comment{The heap metadata is stored on a single node. Currently all heap
allocation and free operations are serialized through this node; while this
has been sufficient for our benchmarks, it is easy to relax in the future
\Grappa will provide parallel performance through
combining~\cite{MAMA,flatcombining}.}

\Grappa also allows any local data on a core's stacks or heap to be
exported to the global address space to be made accessible to other
cores across the system. Addresses to global memory allocated in this
way use \textbf{2D global addresses}.  This uses a traditional PGAS
addressing model, where each address is a tuple of a rank in the job (or
global process ID) and an address in that process. The lower 48 bits of
the address hold a virtual address in the process. The top bit is set to
indicates that the reference is a 2D address (as opposed to linear
address). This leaves 15 bits for network endpoint ID, which limits our
scalability to $2^{15}$ endpoints. Any node-local data can be made
accessible by other nodes in the system by wrapping the address and node
ID into a 2D global address. This address can then be accessed with a
delegate operation and even be cached by other nodes. The
address is converted at the destination into a canonical x86 address by replacing the upper
bits with the sign-extended upper bit of the virtual address. 2D
addresses may refer to memory allocated from a single processes' heap or
from a task's stack. Figure~\ref{fig:memory-structure} shows how 2D and
linear addresses can refer to other cores' memory.

\begin{figure}[htbp]
  \begin{center}
    \begin{minipage}{\columnwidth}
	\small
	\begin{tabular}{l}
      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
      	\texttt{\scriptsize global\_free( global\_address )} \\ 
      	Allocation in the global heap \\ \\
      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_read\_async( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_write\_async( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ 
      	Memory operations at the home core of a global address \\ \\
      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ 
		Perform cache operations to acquire/release global data.  \\
		Acquire copies all data to local node and returns a pointer. \\ 	
		Release optionally copies data back to global memory. \\
	\end{tabular}
      \caption{\label{fig:accessing-memory} \Grappa API for memory accesses. \TODO{Update table with new calls. And async calls for delegates.}}     \end{minipage}
  \end{center}
\end{figure}

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/memory-structure}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:memory-structure} Global memory referencing in \Grappa}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\paragraph{Global memory access} There are two general methods \Grappa
applications use to {\em access} global memory. One method is to use {\em
delegate} operations, which are short memory accesses performed at the memory
location's home node. This method is used when there is no locality in the
access. The other method is to do {\em explicit caching} of memory regions,
used when the programmer expects that the associated computation has data
locality. Below we explain each in detail.

@@

Delegate operations can be either {\em synchronous} or {\em asynchronous}.
With synchronous operations, the task issuing the delegate call blocks until
the delegate operation completes, which is necessary, for example, to ensure
that synchronization has finished before continuing. Synchronization in \Grappa
is implemented using synchronous delegate operations. Often remote data
accesses do not need to be performed synchronously. In order to avoid
unnecessary waiting, we support asynchronous delegate operations. For reads,
we support a ``futures''-like mechanism. Tasks may issue delegate reads in
parallel and block on the ``promises'' returned by asynchronous delegate
invocations. Delegate write operations may also be performed asynchronously,
and since they do not return data, they need a mechanism to detect completion.
\Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
outstanding asynchronous operations of the local core. The code and can probe
the state of this object to determine when all outstanding requests are
complete. This is necessary to, say, make sure all outstanding operations are
complete before executing a synchronization operation.

@@

\vspace{1ex}
\textit{Explicit caching.} \Grappa provides an API to fetch a global pointer of
any length and return a local pointer to a cached copy of the global memory.
\Grappa cache operations have the usual read-only and read-write variants,
along with a write-only variant used to initialize data structures. Languages
for distributed shared memory systems have done optimizations to achieve a
similar goal. For example, the UPC compiler coalesces struct and array
accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's
message vectorization hoists small messages out of a loop
\cite{FortranD:1992}. Caching in \Grappa additionally provides a mechanism for
exploiting temporal locality when available by operating on the data locally.

Under the hood, \Grappa performs the mechanics of gathering chunks of data from
multiple system nodes and presenting a conventional appearing linear block of
memory as a local pointer into a cache. The strategy employed is to issue all
the constituent requests of the desired block (as Active Messages) and then
yield until all responses have occurred. Currently, \Grappa caches are
\emph{not\/} kept coherent automatically. If that is desired, the programmer is
responsible for ensuring sufficient synchronization (via delegate operations)
occurs before acquiring or releasing cache objects.
